A game-theory modeling approach to utility and strength of interactions dynamics in biomedical research social networks

### 2017-03-29 casm.tex
Introduction updated, and new references to bib file

The analysis of cooperation in scientific research has been the subject of a
number of studies
\cite{VermeulenPenders:2013,Newman:2001,Newman:2004,Elango-etal:2012,HernandezLemus:2013,Strasser:2006,Strasser:2012}. This
is not surprising since cooperation and competition are quite important in
today's academic success. How does collaboration happen within a competitive
academic environment and what kind of payoff is present in these settings were
questions considered recently by Wardil and Hauert \cite{Wardil-etal:2015} in the
context of cooperation in multi-authored publications. Also, the role of game
theory over complex scientific information and collaboration networks has
attracted attention, mainly focusing on how long-term strategies may shape
different scenarios for Nash equilibria \cite{Hanauske-etal:2012}. Prisoner's Dilemma
has been used in the study of impact factor and collaboration
\cite{Hara-etal:2003,LiebermanNowak:2005}. \\

Even with all these research efforts, cooperation in the context of
scientific collaboration is still loosely defined and the long term
dynamics of academic cooperation (and its consequences) are yet to be
fully elucidated. Furthermore, to our current knowledge, there has
been no use of game theory and complex network analysis for
understanding how the topology of scientific collaboration networks
affects access to resources among individuals present in the network
\footnote{For an account of scientific collaboration and definitions,
  please refer to \cite{Sonnenwald:2007}.}. Our work aims to contribute
to our current understanding on the matter, especially when agents have
to maximize their access to resources while taking care of their collaboration links.\\

 We explore the network effect on the distribution
   of players having access to certain amount of resources shared by other players in the
   network and the distribution of the strength of interactions among
   them. Implementing games on networks is an active area of
research aimed to understand the evolution of cooperation in networks populated
by selfish agents \cite{SzaboFath:2007,Nowak-etal:92,OshtukiNowak:2006,Santos-etal:2005,Santos-etal:2006}. In many network models in which some game theory games are simulated, agents' decision to cooperate
or defect depend on a specific strategy, such as the well known
\textit{tit-for-tat} \cite{Axelrod:2006,Nowak-etal:2011}. In some other cases, agents
can modify the weight of the interactions with their neighbors
\cite{Santos-etal:2006}. From a different perspective others have explored the effect
of different topologies on the emergence of cooperation
\cite{Santos-etal:2005,HauertDoebeli:2004}.

   Particularly, we implemented two games played simultaneously: 
   one for maximizing individual utility based on the iterated Prisoner's
   Dilemma; the other, a coordination game for maximizing the connection strength between
   players. We are interested in how they affect each other in 
   the context of a network of scientific collaboration under the idea that while
   researchers are interested in maximizing their utilities, they also know that
   it is important to invest in building collaborative relationships. These two
   behaviors are explored in a biomedical research community in M\'exico.\\

In the context of our model, utility represents
  access to resources shared by others.
  The value of the utility function for a player is the sum of the payoffs of playing
  prisoner\'s dilemma with its neighbors. The the other concurrent game consists of players
  trying to maximize the strength of their interactions with other players. In the coordination game the best strategy is to adopt the same strategy as the other player, as it pays the most,
  regardless of cooperation or defection in the utility game. When
  both cooperate the interaction gets a positive payoff, when both defect, the
  interaction doesn't get affected; but if they do not coordinate, then the
  interaction loses. The payoffs resulting from the coordination games are not part of the utility function, however, both are considered by the agent to decide whether to cooperate or defect. \\

@@

\section*{Acknowledgments}
    We acknowledge the help of Francisco Allende for his work curating 
databases of researchers. We acknowledge Be Luecke for her critical review.

>>>>

The analysis of cooperation in scientific research has been the subject of a
number of studies
\cite{VermeulenPenders:2013,Newman:2001,Newman:2004,Elango-etal:2012,HernandezLemus:2013,Strasser:2006,Strasser:2012}. This
is not surprising since cooperation and competition are quite important in
today's academic success. Moreover, the quantitative analysis of cooperation in
science also has a long history of more than 50 years
\cite{Mali-etal:2012,Scharnhorst-etal:2012}. Particularly appealing to researchers studying
this topic has been the use of network analysis, especially in the area of
bibliometrics and scientometrics. The approach we follow here articulates game theory
together with complex network analysis. Putting together game-theoretic modeling
on complex networks is common in studies about the emergence of cooperation
\cite{SzaboFath:2007,Nowak-etal:92,OshtukiNowak:2006,Santos-etal:2005,Santos-etal:2006}.
However, in the context of science studies, it is far from being a unified, active
field of research. There is no common set of questions and problems or a
coherent body of models and theory aimed to understand the quirks and twists of
cooperation in scientific communities. It is more likely to say that there is a
desarticulated collection of works, implementing game theory and
complex networks to different situations related to the social life of science.\\

Nevertheless, there is work done. For instance Wardil and Hauert have analyzed
how does collaboration happen within co-authorship networks
\cite{Wardil-etal:2015}. Using the \emph{Physical Review} journal historic database,
these authors show the payoff differences between being the first author and the
last author, where the first author is a cooperator that takes the highest load
of work, while the last author is the \emph{laird} that can benefit from the
work of the former and get engaged in many other projects. They argue that
cooperation in a multi-authored paper resambles more to a snow-drift game than a
prisoner's dilemma situation, since when co-authoring a paper, every body wins
but each author pays a different price. Though this work takes into account the
topology of the network as informative regarding cooperation patterns, the role
of topology framing or
fostering cooperation is not part of the study. \\

Game theory over complex scientific information and collaboration networks has
also attracted attention, for instance Hanauske has focused on how long-term
strategies may shape different scenarios for Nash equilibria in the interaction
between scientists and journals \cite{Hanauske-etal:2012}. And Hara and
collaborators have used Prisoner's Dilemma in the study of impact factor and
collaboration \cite{Hara-etal:2003,LiebermanNowak:2005}. \\

% In general and beyond the area of scientific collaboration,
% implementing games on networks is an active area of research aimed to understand
% the evolution of cooperation in networks populated by selfish agents
% \cite{SzaboFath:2007,Nowak-etal:92,OshtukiNowak:2006,Santos-etal:2005,Santos-etal:2006},
% for example, by exploring different games such as the prisoner's dilemma,
% snow-drift, etc., \cite{Axelrod:2006,Nowak-etal:2011}, or implementing models in
% which agents can modify the weight of the interactions with their neighbors
% \cite{Santos-etal:2006}. \\

Even with all these research efforts, cooperation in the context of scientific
collaboration is still loosely defined and the long term dynamics of academic
cooperation (and its consequences) are yet to be fully elucidated. Furthermore,
to our current knowledge, there has been no use of game theory and complex
network analysis for understanding how the topology of scientific collaboration
networks affects access to resources among individuals present in the network
\footnote{For an account of scientific collaboration and definitions, please
  refer to \cite{Sonnenwald:2007}.}. From a different perspective others have
explored the effect of different topologies on the emergence of cooperation
\cite{Santos-etal:2005,HauertDoebeli:2004}. Our work aims to contribute to our
current understanding on the matter, especially when agents have to maximize
their access to resources while taking care of their collaboration links. We
explore the network effect on the distribution of players having access to
certain amount of resources shared by other players in the network and the
distribution of the strength of interactions among them. Particularly, we
implemented two games played simultaneously: one for maximizing individual
utility based on the iterated Prisoner's Dilemma; the other, a coordination game
for maximizing the connection strength between players. We are interested in how
they affect each other in the context of a network of scientific collaboration
under the idea that while researchers are interested in maximizing their
utilities, they also know that it is important to invest in building
collaborative relationships. These two behaviors are explored in a biomedical
research community in M\'exico.\\

   In the context of our model, utility represents access to resources shared by
   others.  The value of the utility function for a player is the sum of the
   payoffs of playing prisoner's dilemma with its neighbors. The other
   concurrent game consists of players trying to maximize the strength of their
   interactions with other players. In the coordination game the best strategy
   is to adopt the same strategy as the other player, as it pays the most,
   regardless of cooperation or defection in the utility game. When both
   cooperate the interaction gets a positive payoff, when both defect, the
   interaction doesn't get affected; but if they do not coordinate, then the
   interaction loses. The payoffs resulting from the coordination games are not
   part of the utility function, however, both are considered by the agent to
   decide whether to cooperate or defect. \\

@@

\section*{Acknowledgments}
    We acknowledge the help of Ma. del Pilar Galarza and Francisco Allende for their work curating 
databases of researchers. We acknowledge Be Luecke for her critical review.
